---
title: "EDA and Analysis"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#load required packages
```{r}
library(tidyverse)
library(ggplot2)
library(caret)
library(mgcv)
library(ggthemes)
library(car)
library(mixexp)
library(MASS)
```

#load data
```{r}
df <- as.data.frame(X_dffinal)
head(df)
```

#predictor variable (regular season win %) analysis
```{r}
ggplot(df, aes(x=rswp)) +
        geom_histogram(aes(y=..density..), bins=15, fill="steel blue",
                       color = "white") +
        geom_density(color="red") + theme_few() +
        ggtitle("Density Histogram of Regular Season Win Percentage")

ggplot(df, aes(sample=rswp)) +
        stat_qq(aes(sample=rswp), color="steel blue")+ 
        theme_few()+
        ggtitle("RSWP Quantile Plot")
```
# Positional cap share analysis
```{r}
#create linear models to find any correlation
ggplot(df, aes(x=qb_cap_share)) +
        geom_histogram(bins=15, fill="steel blue",
                       color = "white") + theme_few() +
        ggtitle("Histogram of QB Proportion of Total Spend")
#data appears approximately normally distributed
ggplot(df, aes(x=wr_cap_share)) +
        geom_histogram(bins=15, fill="steel blue",
                       color = "white") + theme_few() +
        ggtitle("Histogram of WR Proportion of Total Spend")
#data appears approximately normally distributed
hist(df$wr_cap_share)
lm1 <- lm(rswp ~ qb_cap_share + rb_cap_share + wr_cap_share + te_cap_share+
                           ol_cap_share + dl_cap_share + lb_cap_share + sec_cap_share, data = df) 
summary(lm1)

mix_df <- data.frame(df$qb_cap_share, df$rb_cap_share, df$wr_cap_share, df$te_cap_share, df$ol_cap_share, df$dl_cap_share, df$lb_cap_share, df$sec_cap_share, df$st_cap_share, df$rswp)
mix_capshare_model <- MixModel(mix_df, "df.rswp", mixcomps=c("df.qb_cap_share", "df.rb_cap_share", "df.wr_cap_share", "df.te_cap_share", "df.ol_cap_share", "df.dl_cap_share", "df.lb_cap_share", "df.sec_cap_share", "df.st_cap_share"), model = 4)
summary(mix_capshare_model)

```

# Draft Allocation Variable Analysis
```{r}
ggplot(df, aes(x=qb_pa)) +
        geom_histogram(bins=15, fill="steel blue",
                       color = "white") + theme_few() +
        ggtitle("Histogram of QB Draft Capital Proportion")
#data appears approximately normally distributed
ggplot(df, aes(x=lb_pa)) +
        geom_histogram(bins=15, fill="steel blue",
                       color = "white") + theme_few() +
        ggtitle("Histogram of LB Draft Capital Proportion")
symbox(df$qb_pa) 
# log transformation looks the most appropriate
# since other histograms look similar and for consistency's sake will use same transform for each variable

lm2 <- lm(rswp ~ log(qb_pa+.01) + log(rb_pa+.01) + log(wr_pa+.01) + log(te_pa+.01) +
                       log(ol_pa+.01) + log(dl_pa+.01) + log(lb_pa+.01) + log(sec_pa+.01),
                       data = df)
summary(lm2)
#Plot linebacker variable
ggplot(df, aes(x=log(lb_pa+.01), y=rswp))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_smooth(method="lm", color="red")
```


#Remaining variable analysis
#Normalized Total Spend
```{r}
ggplot(df, aes(x=norm_total_spend)) +
        geom_histogram(aes(y=..density..), bins=15, fill="steel blue",
                       color = "white") +
        geom_density(color="red") + theme_few() +
        ggtitle("Density Histogram of Normalized total Cap Spend")
# data appears to be approximately normally distributed

ggplot(df, aes(x=norm_total_spend, y=rswp))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_smooth(method="lm", color="red")

```

# Analysis of extension vs fa signing
```{r}
#analyze norm fa spend variable
ggplot(df, aes(x=norm_fa_spend)) +
        geom_histogram(bins=15, fill="steel blue",
                       color = "white") + theme_few() +
        ggtitle("Histogram of Normalized FA Spend")
symbox(df$norm_fa_spend) #appears we should use a sqrt or log transform
ggplot(df, aes(x=norm_fa_spend^.5)) +
        geom_histogram(bins=15, fill="steel blue",
                       color = "white") + theme_few() +
        ggtitle("Histogram of Normalized FA Spend ^.5")

ggplot(df, aes(x=norm_fa_spend^.5, y=rswp))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_smooth(method="lm", color="red")

#turn ext rank into categorical variable with 4 levels
df$ext_spend_rank <- as.character(df$ext_spend_rank)
df$ext_spend_rank[df$ext_spend_rank %in% c("1","2", "3", "4", "5", "6", "7", "8")] <- "1"
df$ext_spend_rank[df$ext_spend_rank %in% c("9","10", "11", "12", "13", "14", "15", "16")] <- "2"
df$ext_spend_rank[df$ext_spend_rank %in% c("17","18", "19", "20", "21", "22", "23", "24")] <- "3"
df$ext_spend_rank[df$ext_spend_rank %in% c("25","26", "27", "28", "29", "30", "31", "32")] <- "4"
df$ext_spend_rank <- as.factor(df$ext_spend_rank)

```

# Analysis age
```{r}
ggplot(df, aes(x=avg_age)) +
        geom_histogram(bins=30, fill="steel blue",
                       color = "white") + theme_few() +
        ggtitle("Histogram of Average Age")

ggplot(df, aes(x=avg_age, y=rswp))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_smooth(method="lm", color="red")
```
# Trades Analysis
```{r}
hist(df$traded_to) # appears a transform is needed
symbox(df$traded_to) #use sqrt transform

ggplot(df, aes(x=traded_to^.5)) +
        geom_histogram(bins=15, fill="steel blue",
                       color = "white") + theme_few() +
        ggtitle("Histogram of Traded_to^.5")

ggplot(df, aes(x=traded_to^.5, y=rswp))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_smooth(method="lm", color="red")

hist(df$traded_from) # appears a transform is needed
symbox(df$traded_from) # use sqrt transfrom

ggplot(df, aes(x=traded_from^.5)) +
        geom_histogram(bins=15, fill="steel blue",
                       color = "white") + theme_few() +
        ggtitle("Histogram of Traded_from^.5")

ggplot(df, aes(x=traded_from^.5, y=rswp))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_smooth(method="lm", color="red")
```

#Linear Model of Remaining Variables 
```{r}
lm3 <- lm(rswp ~ norm_total_spend + I(norm_fa_spend^.5) + ext_spend_rank + avg_age +
                  I(traded_to^.5) + I(traded_from^.5), data = df)
summary(lm3) 
```
Final linear Model
```{r}
lm_final <- lm(rswp ~ norm_total_spend + log(lb_pa + .01)  + I(norm_fa_spend^.5) + 
                  I(traded_to^.5) + avg_age, data = df)
summary(lm_final)

lm_final <- lm(rswp ~ norm_total_spend + log(lb_pa + .01)  + I(norm_fa_spend^.5) + 
                  I(traded_to^.5), data = df)
summary(lm_final)

```
# Outliers
```{r}
#Identify high leverage points for each model
influenceIndexPlot(lm_final, vars = c("Studentized"))
influenceIndexPlot(lm_final)


#After looking into observations 254 and 261 there is no compelling reason to remove these observations from the dataset
df_mod <- df[-c(254,261), ]
lm_adjusted <- lm(rswp ~ norm_total_spend + I(norm_fa_spend^.5) +
                  I(traded_to^.5) + log(lb_pa + .01), data = df_mod)
summary(lm_adjusted)

#excluding the cook's high leverage points is actually adding R^2 into model....there is no compelling reason to exclude the data
```
#GAM
# First kernel smoothing
```{r}
sd(df$rswp) # clculate null model rswp = .1919

# begin w/ norm total spend kernel smooth
c1 <- ksmooth(df$norm_total_spend, df$rswp, bandwidth = bw.nrd(df$norm_total_spend),
              x.points = df$norm_total_spend) 
fit1 <- rep(0, 288)
fit1[order(df$norm_total_spend)] <- c1$y

#find confidence intervals
fm1 <- matrix(data = NA, nrow = 200, ncol = 100)
for(i in 1:200){
        samp_index <- sample(1:288, 100, replace = T)
        x <- df[samp_index, 14]
        y <- df[samp_index, 3]
        c <- ksmooth(x, y, bandwidth = bw.nrd(df$norm_total_spend), x.points = x)
        fm1[i, ] <- c$y
}

fm1_sorted <- matrix(data=NA, nrow = 200, ncol = 100)
for(i in 1:100){
        fm1_sorted[, i] <- sort(fm1[, i])
}

system("R CMD SHLIB conf.c")
dyn.load("conf.so")
n <- 200
m <- 100
z <- .C("lowp", as.double(fm1_sorted), as.integer(n), as.integer(m), as.double(y))
z1 <- .C("highp", as.double(fm1_sorted), as.integer(n), as.integer(m), as.double(y))
low_percentile <- z[[4]]
high_percentile <- z1[[4]]
g1 <- seq(min(df_mod$norm_total_spend), max(df_mod$norm_total_spend), length = 100)
df_pct_1 <- data.frame(low_percentile, high_percentile, g1)

#Plot Kernel Density w/ conf intervals
#test plot
ggplot(df, aes(x=norm_total_spend, y=rswp))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_line(aes(df$norm_total_spend, fit1, color="red"))+
        ggtitle("First Kernel Smoothing")+
        geom_line(data=df_pct_1, aes(y=low_percentile, x=g1, color="black"))+
        geom_line(data=df_pct_1, aes(y=high_percentile, x=g1, color="black"))+
        theme(legend.position="none")


y1 <- df$rswp - fit1
rmse1 <- (sum(y1^2)/288)^.5
x2 <- log(df$lb_pa+.01)
df2 <- data.frame(x2, y1)
ggplot(data=df2, aes(x=x2, y=y1))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        ggtitle("Residuals after First Kernel Smoothing")+
        xlab("Log of % Draft Allocation to Linebacker Position +.01")
```

#second kernel smoothing
```{r}
c2 <- ksmooth(df2$x2, df2$y1, bandwidth = bw.nrd(df2$x2), x.points = df2$x2)
fit2 <- fit1
fit2[order(x2)] <- c2$y

#find confidence intervals
fm2 <- matrix(data = NA, nrow = 200, ncol = 100)
for(i in 1:200){
        samp_index <- sample(1:284, 100, replace = T)
        x <- df2[samp_index, 1]
        y <- df2[samp_index, 2]
        c <- ksmooth(x, y, bandwidth = bw.nrd(df2$x2), x.points = x)
        fm2[i, ] <- c$y
}

fm2_sorted <- matrix(data=NA, nrow = 200, ncol = 100)
for(i in 1:100){
        fm2_sorted[, i] <- sort(fm2[, i])
}

system("R CMD SHLIB conf.c")
dyn.load("conf.so")
n <- 200
m <- 100
z <- .C("lowp", as.double(fm2_sorted), as.integer(n), as.integer(m), as.double(y))
z1 <- .C("highp", as.double(fm2_sorted), as.integer(n), as.integer(m), as.double(y))
low_percentile2 <- z[[4]]
high_percentile2 <- z1[[4]]
g2 <- seq(min(x2), max(x2), length = 100)
df_pct_2 <- data.frame(low_percentile2, high_percentile2, g2)

#Plot Kernel Density w/ conf intervals
ggplot(df2, aes(x=x2, y=y1))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_line(aes(df2$x2, fit2, color="red"))+
        ggtitle("Second Kernel Smoothing")+
        xlab("Log LB % Draft Capital Allocation + .01")+
        ylab("Residuals after first kernel smoothing")+
        geom_line(data=df_pct_2, aes(y=low_percentile2, x=g2, color="black"))+
        geom_line(data=df_pct_2, aes(y=high_percentile2, x=g2, color="black"))+
        theme(legend.position="none")
        


y2 <- y1- fit2
rmse2 <- (sum(y2^2)/288)^.5 # value of .169
x3 <- (df$norm_fa_spend)^.5
df3 <- data.frame(x3, y2)
ggplot(data=df3, aes(x=x3, y=y2))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        ggtitle("Residuals after Second Kernel Smoothing")+
        xlab("Sqrt Normalized FA Spend")
```

#third kernel smoothing
```{r}
c3 <- ksmooth(df3$x3, df3$y2, bandwidth = bw.nrd(df3$x3), x.points = df3$x3)
fit3 <- fit2
fit3[order(x3)] <- c3$y

#find confidence intervals
fm3 <- matrix(data = NA, nrow = 200, ncol = 100)
for(i in 1:200){
        samp_index <- sample(1:284, 100, replace = T)
        x <- df3[samp_index, 1]
        y <- df3[samp_index, 2]
        c <- ksmooth(x, y, bandwidth = bw.nrd(df3$x3), x.points = x)
        fm3[i, ] <- c$y
}

fm3_sorted <- matrix(data=NA, nrow = 200, ncol = 100)
for(i in 1:100){
        fm3_sorted[, i] <- sort(fm3[, i])
}

system("R CMD SHLIB conf.c")
dyn.load("conf.so")
n <- 200
m <- 100
z <- .C("lowp", as.double(fm3_sorted), as.integer(n), as.integer(m), as.double(y))
z1 <- .C("highp", as.double(fm3_sorted), as.integer(n), as.integer(m), as.double(y))
low_percentile3 <- z[[4]]
high_percentile3 <- z1[[4]]
g3 <- seq(min(x3), max(x3), length = 100)
df_pct_3 <- data.frame(low_percentile3, high_percentile3, g3)

#Plot Kernel Density w/ conf intervals
ggplot(df3, aes(x=x3, y=y2))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_line(aes(df3$x3, fit3, color="red"))+
        ggtitle("Third Kernel Smoothing")+
        xlab("Sqrt Normalized FA Spend")+
        ylab("Residuals after second kernel smoothing")+
        geom_line(data=df_pct_3, aes(y=low_percentile3, x=g3, color="black"))+
        geom_line(data=df_pct_3, aes(y=high_percentile3, x=g3, color="black"))+
        theme(legend.position="none")

y3 <- y2- fit3
rmse3 <- (sum(y3^2)/288)^.5 # value of .166
x4 <- (df$traded_to^.5)
df4 <- data.frame(x4, y3)
ggplot(data=df4, aes(x=x4, y=y3))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        ggtitle("Residuals after Third Kernel Smoothing")+
        xlab("Sqrt of Players Traded To")

```

# Fourth Kernel Smoothing
```{r}
c4 <- ksmooth(df4$x4, df4$y3, bandwidth = bw.nrd(df4$x4), x.points = df4$x4)
fit4 <- fit3
fit4[order(x4)] <- c4$y

#find confidence intervals
fm4 <- matrix(data = NA, nrow = 200, ncol = 100)
for(i in 1:200){
        samp_index <- sample(1:284, 100, replace = T)
        x <- df4[samp_index, 1]
        y <- df4[samp_index, 2]
        c <- ksmooth(x, y, bandwidth = bw.nrd(df4$x4), x.points = x)
        fm4[i, ] <- c$y
}

fm4_sorted <- matrix(data=NA, nrow = 200, ncol = 100)
for(i in 1:100){
        fm4_sorted[, i] <- sort(fm4[, i])
}

system("R CMD SHLIB conf.c")
dyn.load("conf.so")
n <- 200
m <- 100
z <- .C("lowp", as.double(fm4_sorted), as.integer(n), as.integer(m), as.double(y))
z1 <- .C("highp", as.double(fm4_sorted), as.integer(n), as.integer(m), as.double(y))
low_percentile4 <- z[[4]]
high_percentile4 <- z1[[4]]
g4 <- seq(min(x4), max(x4), length = 100)
df_pct_4 <- data.frame(low_percentile4, high_percentile4, g4)

ggplot(df4, aes(x=x4, y=y3))+
        geom_point(color="steel blue",cex=.7)+theme_few()+
        geom_line(aes(df4$x4, fit4, color="red"))+
        ggtitle("Fourth Kernel Smoothing")+
        xlab("Sqrt of # Players Traded To")+
        ylab("Residuals after third kernel smoothing")+
        geom_line(data=df_pct_4, aes(y=low_percentile4, x=g4, color="black"))+
        geom_line(data=df_pct_4, aes(y=high_percentile4, x=g4, color="black"))+
        theme(legend.position="none")

y4 <- y3- fit4
rmse4 <- (sum(y4^2)/288)^.5 #.1609
mod_r_sqr <- 1- rmse4/sd(df$rswp) #.1616 which is slightly worse than linear R^2

#add hist of residuals

ggplot(data.frame(y4), aes(x=y4)) +
        geom_histogram(bins=20, fill="steel blue", color = "white") +
        theme_few() + ggtitle("Residuals after 4th Kernel Smoothing")+
        theme(legend.position="none")

ggplot(data.frame(y4), aes(sample=y4)) +
        stat_qq(aes(sample=y4), color="steel blue")+ 
        theme_few()+
        ggtitle("Residual Quantile Plot")

        

```
# MISC PLots
```{r}
means_cs <- c(mean(df$qb_cap_share), mean(df$rb_cap_share), mean(df$wr_cap_share), mean(df$te_cap_share),
                     mean(df$ol_cap_share), mean(df$dl_cap_share), mean(df$lb_cap_share), mean(df$sec_cap_share),
                     mean(df$st_cap_share))
means_drft <- c(mean(df$qb_pa), mean(df$rb_pa), mean(df$wr_pa), mean(df$te_pa),
                     mean(df$ol_pa), mean(df$dl_pa), mean(df$lb_pa), mean(df$sec_pa),
                     mean(df$st_pa))

mean_vec <- c(means_cs, means_drft)
type_vec <- rep(c("cap", "draft"), each=9)
pos_vec <- rep(c("QB", "RB", "WR", "TE", "OL", "DL", "LB", "SEC", "ST"), 2)
df_means <- data.frame(pos_vec, type_vec, mean_vec)
colnames(df_means) <- c("position", "type", "mean")

ggplot(data=df_means, aes(position, mean, fill=type))+
        geom_bar(stat = "identity", position = 'dodge')+theme_few()+
        ggtitle("Mean Cap and Draft Allocation by Position")+
        scale_fill_fivethirtyeight()
                
        
        
```





